The Truth is Revealed: Identity of the Fish Fry "Teti" Fishery in Eastern Cuba

### 2020-11-30 analysis/paper/paper.Rmd
paper & bib

---
title: 'The Truth is Revealed: Identity of the Fish Fry "Teti" Fishery in Eastern Cuba'
author:
  - Sheila Rodriguez-Machado:
      email: srodr62@lsu.edu
      institute: [UofO]
      correspondence: true
  - Tomas M. Rodriguez-Cabrera:
      email: tomasmichelrodriguez@gmail.com
      institute: [UofA]
      correspondence: false
institute:
  - UofO: Museum of Natural Sciences, Louisiana State University
  - UofA: Sociedad Cubana de Zoologia
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: '/Users/sheila_ffish/Documents/rrcompendium/analysis/templates/template.docx' # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: '/Users/sheila_ffish/Documents/rrcompendium/analysis/templates/journal-of-archaeological-science.csl' # Insert path for the bib-style
abstract: |
  texting texting
keywords: |
  amphidromy; Busck's Stone-Biting Goby; COI; Gobiidae

@@

#knitr::opts_knit$set(root.dir = '../rrcompendium/')

#getwd()

@@

# INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("raster", "ggplot2", "ggthemes",  
             "rentrez", "ape", "seqinr", "msa", "Biostrings", "systemPipeR", "phangorn", "ggtree")

@@

@@

Tetis (19–28 mm standard length) were collected 400 m from the river mouth of Río Yumurí (Figure 2.1), eastern Cuba, in December, 2018. Samples were collected using hand nets (1 mm mesh size) 2 m from the riverbank. Specimens were anesthetized using MS 222 (Tricaine Methanesulphonate) and immediately preserved in 80% ethanol. Voucher specimens were deposited in the collection of the Acuario Nacional de Cuba (catalog number: ANC-xxxx).

@@

PCR products were purified using ExoSAP-IT™ PCR Product Cleanup Reagent (Affymetrix Inc./ USB, Cleveland, OH, USA), and then sequenced on an ABI 3100 automated sequencer (Applied Biosystems). Raw sequence files were edited and aligned with the software Geneious v.9.1 [@kearse2012geneious], and submitted to GenBank database (accession numbers #####). We used sequences of the genus *Sicydium* as reference for recovering the taxonomic identity of the "tetis". Sequences of *Awaous banana* and *A. ocellaris* were used as outgroup. All non-Cuban sequences were recovered from GenBank using the R package *rentrez* [@winter2017rentrez]. Duplicated sequences and sequences with missing data were excluded. Sequences were edited using the R packages *seqinr* [@charif2007seqinr], *ape* [@paradis2015package], and *biostrings* [@pages2013package]. Final alignment was obtained with the algorithm ClustalOmega (@sievers2011fast) using the R package *msa* [@bodenhofer2015msa].

@@

# storing ALL sequences in a single fasta file
write(Sicydium_COI_seqs1, file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_1-seqs.fasta", sep = "\n")
write(Sicydium_COI_seqs2, file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_1-seqs.fasta", sep = "\n", append = TRUE)   # 'append = TRUE', I can keep adding seqs
write(Awaous_bananaCOI_seqs, file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_1-seqs.fasta", sep = "\n", append = TRUE)
write(Awaous_ocellarisCOI_seqs, file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_1-seqs.fasta", sep = "\n", append = TRUE)

@@

Sicydium_COI_seqinrFormat <- read.fasta(file ="../rrcompendium/analysis/data/raw_data/Sicydium_COI_1-seqs.fasta", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

@@

# writing a new fasta file (2nd) using the vector above 
write.fasta(sequences = Sicydium_COI_seqinrFormat, names = Sicydium_COI_GenBank_Ids, nbchar = 10,
            file.out = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta")

@@

### REMOVING sequences of incomplete/unspecific NAMES; that is, those with 'sp.'
nameFasta <- read.fasta(file = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta",
                        seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

@@

# writing a new fasta (3rd) file with only those seqs with full scientific names
write.fasta(sequences = nameFasta[completeSeqs], names = names(nameFasta)[completeSeqs],
            file.out = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_3-names.fasta")

### REMOVING sequences of LENGTH shorter than 600 pb, given that the gene COI has between 627 and 658 bp.
lengthFasta <- read.fasta(file = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_3-names.fasta",
                          seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)                     # 586 seqs
fastaLength <- lengthFasta[which(getLength(lengthFasta) >= 600)]
write.fasta(sequences = fastaLength, names = names(fastaLength),
            file.out = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_4-lengths.fasta")             # 401 seqs

### Removing DUPLICATE sequences
# using "seqinr" to do it 
# for not using attributes to name sequences
duplicatesFasta <- read.fasta("../rrcompendium/analysis/data/raw_data/Sicydium_COI_4-lengths.fasta", set.attributes = FALSE)

seqnames <- names(duplicatesFasta)
dups <- grep(TRUE,duplicated(duplicatesFasta))
uniqueNames <- seqnames[-dups]
uniqueSeqs <- duplicatesFasta[-dups] 
write.fasta(sequences = uniqueSeqs, names = uniqueNames,
            file.out = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_5-dupli.fasta")               # 204 seqs

# adding samples from Cuba (n=6)
mySamples <- read.fasta("../rrcompendium/analysis/data/raw_data/Sicydium_COI_Cuba.fas")
write.fasta(mySamples, names(mySamples),
            "../rrcompendium/analysis/data/raw_data/Sicydium_COI_5-dupli.fasta", open = "a")              # 210 seqs

@@

alignmentFasta <- readDNAStringSet("../rrcompendium/analysis/data/raw_data/Sicydium_COI_5-dupli.fasta")
ClustalOAlignment <- msa(alignmentFasta, "ClustalOmega")
# print(ClustalOAlignment, showConsensus=FALSE, showNames=TRUE, show="complete")

# Checking for the position of sequences that were not edited properly so they still have nucleotides different that A, G, C, T 
nucType <- alphabetFrequency(ClustalOAlignment)
unknownNuc <- nucType[, c(5:14)]
unknownPosition <- which(unknownNuc > 0, arr.ind=T)

# REMOVING samples with too many GAPS at the beginning and the end of the sequences and UNKNOWN nucleotides (ex.: W, K, S) 
colmask(ClustalOAlignment) <- IRanges(start=c(1,666),end=c(25,679))
rowmask(ClustalOAlignment) <- IRanges(start=c(1),end=c(1))
ClustalOAlignment2 <- as(ClustalOAlignment, "DNAStringSet")
writeXStringSet(ClustalOAlignment2,
                file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_6-alignment.fasta")           # 209 seqs -- 640 bp

### Replacing gaps at the begining and end with Ns
gapsFasta <- read.fasta(file = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_6-alignment.fasta", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE, apply.mask = FALSE)
#getLength(gapsFasta)
replacingGaps <- lapply(gapsFasta, function(x) gsub("-", "N", x))
write.fasta(replacingGaps, names(replacingGaps), "../rrcompendium/analysis/data/raw_data/Sicydium_COI_7-gaps.fasta")

@@

alignedFasta <- readDNAStringSet("../rrcompendium/analysis/data/raw_data/Sicydium_COI_7-gaps.fasta")

getRF <- function(alignedFasta){
  if (class(alignedFasta)!= c("DNAStringSet")){
    stop("This function only takes files of class 'DNAStringSet'. Use package 'Biostrings' to read the input file")
  }
  if (any(letterFrequency(alignedFasta, "-", as.prob=TRUE))>0){
    stop("This function do not take gaps as values/strings. Go and replaced gaps with Ns (if the gaps are at the begining or end of the sequences) or choose a subset of sequences that do not contain gaps")
  }
  orf1 <- predORF(alignedFasta[1:10], n=1, type="df", mode="orf", strand="sense")   # predORF() do not read seqs with gaps
  # predicts longest ORF for sense strand in each query sequence                                                                          
  longestSeq1 <- orf1$start[orf1$width == max(orf1$width)]
  codonPosition1 <- longestSeq1[1]/3
  codonConfirmation1 <- round(codonPosition1)
  if (codonPosition1==codonConfirmation1){
   file.rename("../rrcompendium/analysis/data/raw_data/Sicydium_COI_7-gaps.fasta", "../rrcompendium/analysis/data/raw_data/Sicydium_COI_newRF.fasta")                     # 640 bp
  print("The Reading Frame of these seqs is +1, no further action is required, go to modelTesting")
    } else {
    if (codonPosition1!=codonConfirmation1){
      tempLength <- length(alignedFasta[[1]])
      #remove first nucleotide position
      newRF2 <- subseq(alignedFasta, start = 2, end = tempLength)      # 639 bp
      # getting new RF from the new alignment with -1 bp.
      orf2 <- predORF(newRF2[1:10], n=1, type="df", mode="orf", strand="sense") 
      # checking size gain
      longestSeq2 <- orf2$start[orf2$width == max(orf2$width)]
      codonPosition2 <- (longestSeq2[1]+2)/3
      codonConfirmation2 <- round(codonPosition2)
      if (codonPosition2==codonConfirmation2){
        writeXStringSet(newRF2, file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_newRF.fasta")
      print("The Reading Frame of these seqs was +2, new fasta file created, go to modelTesting")
      } else {
      if (codonPosition2!=codonConfirmation2){
        #remove first nucleotide position
        tempLength <- length(alignedFasta[[1]])
        newRF3 <- subseq(alignedFasta, start = 3, end = tempLength)
        writeXStringSet(newRF3, file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_newRF.fasta")    # 638 bp
      print("The Reading Frame of these seqs was +3, new fasta file created, go to modelTesting")
     }
   }  
  }
 }
}

@@

The nucleotide substitution model that best explained our data was estimated using the function modelTest from the R package phangorn [@schliep2011phangorn]. The phylogenetic tree was reconstructed following a maximum mikelihood approach [@felsenstein1981evolutionary] using the functions optim.pml and bootstrap.pml, also from the package phangorn. The tree was generated and visualized using the R package ggtree [@yu2017ggtree].

@@

fastaRF <- read.dna(file="../rrcompendium/analysis/data/raw_data/Sicydium_COI_newRF.fasta", format = "fasta")

fastaRF1 <- read.fasta(file = "../rrcompendium/analysis/data/raw_data/Sicydium_COI_6-alignment.fasta", seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE, apply.mask = FALSE)
getLength(fastaRF1)

@@

```{r tree, include=TRUE, results = 'hide', fig.show='hide'}

@@

# avoid to create a tree every time the code runs
#set.seed(2015-07-01)

fit = optim.pml(fitStart, rearrangement = "stochastic", optGamma=TRUE, optInv=FALSE, model="HKY")     # seeking a maximum likelihood solution
set.seed(123)    # getting the same tree topology each time the code runs 
bs = bootstrap.pml(fit, bs=100, optNni=TRUE, multicore=TRUE, mc.cores = 4)

# saving the tree in the folder 'analysis/figures/'
bsTree <- plotBS(midpoint(fit$tree), bs, p = 75, type="p", bs.col = "blue")     # fit$tree is where the optimal tree is stored
optimalTree <- plotBS(bsTree, type = "p", cex = 0.2, font = 1)
# export tree with bootstrap values as node labels
write.tree(optimalTree, file="../rrcompendium/analysis/data/derived_data/Sicydium_ML_tree.tre")

@@

hshdhdhdh

```{r demo-plot1, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

```{r inline-code2, include = TRUE, results='hide'}
# the result wont be displayed in the final document
loglikelihood <- round(-1*(as.numeric(fit[1])), 2)

@@

# Editing and visualizing the tree to represent --- library: ggtree
tree <- read.tree(file="../rrcompendium/analysis/data/derived_data/Sicydium_ML_tree.tre")

# checkinf whether the tree is rooted
rootedTree <- is.rooted(tree)   # package ape. showing the tree root: geom_rootpoint(color="pink", size=5)

# getting the node number for each main clade. This info is important because those #s are the one needed to edit each main clade/species
nodeNumber <- ggtree(tree) + geom_text(aes(label=node), hjust=-.5, col = "blue", size = 1) # looks horrible but it gives the clade number

@@

```{r demo-plot2, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

@@

>>>>

---
title: 'The Truth is Revealed: Identity of the Fish Fry "Teti" Fishery in Eastern Cuba'
author:
  - Sheila Rodriguez-Machado:
      email: srodr62@lsu.edu
      institute: [UofO]
      correspondence: true
  - Tomas M. Rodriguez-Cabrera:
      email: tomasmichelrodriguez@gmail.com
      institute: [UofA]
      correspondence: false
institute:
  - UofO: Museum of Natural Sciences, Louisiana State University
  - UofA: Sociedad Cubana de Zoologia
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: '../templates/template.docx'
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: '../templates/journal-of-archaeological-science.csl'
abstract: |
  The lack of diagnostic characters to distinguish among species and even genera of gobies (Gobiidae) is still a challenge for science in the 21st century. However, the use of molecular tools applied to species-level identification has brought some light into the taxonomy and systematics of these fish taxa. The fish fry fishery in Eastern Cuba known as “Teti” represents a local fishery and there is no information about the exploitation of this natural resource outside this region. Fishing “Teti” has been a common practice for several years but an exponential increase of tourism in Eastern Cuba represents an additional threat to our biodiversity. Besides, if “Teti” refers to a single species or a species complex is still a knowledge gap in Cuba. The present study aims to identify the Fish Fry "Teti" Fishery in Eastern Cuba using molecular markers. We conducted a barcoding analysis comparing Cuban samples of gobies with samples of the genera *Sicydium* and *Awaous* available in GenBank. Here, we provide evidence that Busck´s Stone-Biting Goby, *Sicydium buscki*, is a component of the fish fry “Teti” fishery in Eastern Cuba. However, whether this fishery includes a single species or more than one species will remain as a mystery until more studies are conducted. Although this study represents the first effort toward and effective management, more thorough investigations and identification of the species comprising the “Teti” fishery are needed. It is imperative to include them in local management plans and to avoid overfishing this resource.  
keywords: |
  amphidromy; Busck's Stone-Biting Goby; COI; Gobiidae

@@

#knitr::opts_knit$set(root.dir = '/rrcompendium/')

@@

# INSTALLING PACKAGES AND LIBRARIES - if needed
packages = c("BiocManager", "raster", "ggplot2", "ggthemes", "here",  
             "rentrez", "ape", "seqinr", "phangorn")

@@

packagesBioconductor = c("msa", "Biostrings", "systemPipeR", "ggtree")

package.Bioconductor <- lapply(packagesBioconductor, FUN = function(y) {
    if (!require(y, character.only = TRUE)) {
      BiocManager::install(y, dependencies = TRUE)
      library(y, character.only = TRUE, warn.conflicts = FALSE, quietly = TRUE)
    }
  }
)

#if (!requireNamespace("BiocManager", quietly = TRUE))
  #  install.packages("BiocManager")

#BiocManager::install("msa")

@@

Tetis (19–28 mm standard length) were collected 400 m from the river mouth of Río Yumurí (\@ref(fig:mainMap)), eastern Cuba, in December, 2018. Samples were collected using hand nets (1 mm mesh size) 2 m from the riverbank. Specimens were anesthetized using MS 222 (Tricaine Methanesulphonate) and immediately preserved in 80% ethanol. Voucher specimens were deposited in the collection of the Acuario Nacional de Cuba (catalog number: ANC-xxxx).

@@

PCR products were purified using ExoSAP-IT™ PCR Product Cleanup Reagent (Affymetrix Inc./ USB, Cleveland, OH, USA), and then sequenced on an ABI 3100 automated sequencer (Applied Biosystems). Raw sequence files were edited and aligned with the software Geneious v.9.1 [@kearse2012geneious], and submitted to GenBank database (accession numbers #####). We used sequences of the genus *Sicydium* as reference for recovering the taxonomic identity of the "tetis". Sequences of *Awaous banana* and *A. ocellaris* were used as outgroup. All non-Cuban sequences were recovered from GenBank using the R package *rentrez* [@winter2017rentrez]. Duplicated sequences and sequences with missing data were excluded. Sequences were edited using the R packages *seqinr* [@charif2007seqinr], *ape* [@paradis2015package], and *Biostrings* [@pages2013package]. Final alignment was obtained with the algorithm ClustalOmega [@sievers2011fast] using the R package *msa* [@bodenhofer2015msa].

@@

# storing ALL sequences in a single fasta file
write(Sicydium_COI_seqs1, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n")
# 'append = TRUE', I can keep adding seqs
write(Sicydium_COI_seqs2, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)   
write(Awaous_bananaCOI_seqs, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)
write(Awaous_ocellarisCOI_seqs, here(file="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), sep = "\n", append = TRUE)

@@

Sicydium_COI_seqinrFormat <- read.fasta(here(file ="analysis/data/raw_data/Sicydium_COI_1-seqs.fasta"), seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

@@

# writing a new fasta file (2nd) using the vector above 
write.fasta(sequences = Sicydium_COI_seqinrFormat, names = Sicydium_COI_GenBank_Ids, nbchar = 10,
            here(file.out = "analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta"))

@@

### REMOVING sequences of incomplete/unspecific NAMES; that is, those with 'sp.'
nameFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_2-seqinrFormat.fasta"),
                        seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)

@@

# writing a new fasta (3rd) file with only those seqs with full scientific names
write.fasta(sequences = nameFasta[completeSeqs], names = names(nameFasta)[completeSeqs],
            here(file.out = "analysis/data/raw_data/Sicydium_COI_3-names.fasta"))

### REMOVING sequences of LENGTH shorter than 600 pb, given that the gene COI has between 627 and 658 bp.
lengthFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_3-names.fasta"),
                          seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE)     # 586 seqs
fastaLength <- lengthFasta[which(getLength(lengthFasta) >= 600)]
write.fasta(sequences = fastaLength, names = names(fastaLength),
            here(file.out = "analysis/data/raw_data/Sicydium_COI_4-lengths.fasta"))             # 401 seqs

### Removing DUPLICATE sequences
# using "seqinr" to do it 
# for not using attributes to name sequences
duplicatesFasta <- read.fasta(here("analysis/data/raw_data/Sicydium_COI_4-lengths.fasta"), set.attributes = FALSE)

seqnames <- names(duplicatesFasta)
dups <- grep(TRUE,duplicated(duplicatesFasta))
uniqueNames <- seqnames[-dups]
uniqueSeqs <- duplicatesFasta[-dups] 
write.fasta(sequences = uniqueSeqs, names = uniqueNames,
            here(file.out = "analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"))               # 204 seqs

# adding samples from Cuba (n=6)
mySamples <- read.fasta(here("analysis/data/raw_data/Sicydium_COI_Cuba.fas"))
write.fasta(mySamples, names(mySamples), here("analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"), open = "a")              # 210 seqs

@@

alignmentFasta <- readDNAStringSet(here("analysis/data/raw_data/Sicydium_COI_5-dupli.fasta"))
ClustalOAlignment <- msa(alignmentFasta, "ClustalOmega")
# print(ClustalOAlignment, showConsensus=FALSE, showNames=TRUE, show="complete")

# Checking for the position of sequences that were not edited properly so they still have nucleotides different that A, G, C, T 
nucType <- alphabetFrequency(ClustalOAlignment)
unknownNuc <- nucType[, c(5:14)]  # this indexing only contains nucleotides that are not A, G, T, C, or N or gaps
unknownPosition <- which(unknownNuc > 0, arr.ind=T)

# REMOVING samples with too many GAPS at the beginning and the end of the sequences and UNKNOWN nucleotides (ex.: W, K, S) 
colmask(ClustalOAlignment) <- IRanges(start=c(1,666),end=c(25,679))
rowmask(ClustalOAlignment) <- IRanges(start=c(1),end=c(1))
ClustalOAlignment2 <- as(ClustalOAlignment, "DNAStringSet")
writeXStringSet(ClustalOAlignment2, here(file="analysis/data/raw_data/Sicydium_COI_6-alignment.fasta"))            # 209 seqs -- 640 bp

### Replacing gaps at the begining and end with Ns
gapsFasta <- read.fasta(here(file = "analysis/data/raw_data/Sicydium_COI_6-alignment.fasta"), seqtype = "DNA", as.string = TRUE, forceDNAtolower = FALSE, apply.mask = FALSE)
#getLength(gapsFasta)
replacingGaps <- lapply(gapsFasta, function(x) gsub("-", "N", x))
write.fasta(replacingGaps, names(replacingGaps), here("analysis/data/raw_data/Sicydium_COI_7-gaps.fasta"))

@@

alignedFasta <- readDNAStringSet(here("analysis/data/raw_data/Sicydium_COI_7-gaps.fasta"))

getRF <- function(alignedFasta){
  if (class(alignedFasta)!= c("DNAStringSet")){
    stop("This function only takes files of class 'DNAStringSet'. Use package 'Biostrings' to read the input file")
  }
  if (any(letterFrequency(alignedFasta, "-", as.prob=TRUE))>0){
    stop("This function do not take gaps as values/strings. Go and replaced gaps with Ns (if the gaps are at the begining or end of the sequences) or choose a subset of sequences that do not contain gaps")
  }
  orf1 <- predORF(alignedFasta[1:10], n=1, type="df", mode="orf", strand="sense")   # predORF() do not read seqs with gaps
  # predicts longest ORF for sense strand in each query sequence                                                                          
  longestSeq1 <- orf1$start[orf1$width == max(orf1$width)]
  codonPosition1 <- longestSeq1[1]/3
  codonConfirmation1 <- round(codonPosition1)
  if (codonPosition1==codonConfirmation1){
   file.rename(here("analysis/data/raw_data/Sicydium_COI_7-gaps.fasta"), here("analysis/data/raw_data/Sicydium_COI_newRF.fasta")) 
  print("The Reading Frame of these seqs is +1, no further action is required, go to modelTesting")
    } else {
    if (codonPosition1!=codonConfirmation1){
      tempLength <- length(alignedFasta[[1]])
      #remove first nucleotide position
      newRF2 <- subseq(alignedFasta, start = 2, end = tempLength)      # 639 bp
      # getting new RF from the new alignment with -1 bp.
      orf2 <- predORF(newRF2[1:10], n=1, type="df", mode="orf", strand="sense") 
      # checking size gain
      longestSeq2 <- orf2$start[orf2$width == max(orf2$width)]
      codonPosition2 <- (longestSeq2[1]+2)/3
      codonConfirmation2 <- round(codonPosition2)
      if (codonPosition2==codonConfirmation2){
        writeXStringSet(newRF2, here(file="analysis/data/raw_data/Sicydium_COI_newRF.fasta"))                    
      print("The Reading Frame of these seqs was +2, new fasta file created, go to modelTesting")
      } else {
      if (codonPosition2!=codonConfirmation2){
        #remove first nucleotide position
        tempLength <- length(alignedFasta[[1]])
        newRF3 <- subseq(alignedFasta, start = 3, end = tempLength)
        writeXStringSet(newRF3, here(file="analysis/data/raw_data/Sicydium_COI_newRF.fasta"))                                
      print("The Reading Frame of these seqs was +3, new fasta file created, go to modelTesting")
     }
   }  
  }
 }
}

@@

The nucleotide substitution model that best explained our data was estimated using the function modelTest from the R package *phangorn* [@schliep2011phangorn]. The phylogenetic tree was reconstructed following a maximum likelihood approach [@felsenstein1981evolutionary] using the functions optim.pml and bootstrap.pml, also from the package *phangorn*. The tree was generated and visualized using the R package *ggtree* [@yu2017ggtree].

This paper was completely developed in RStudio, PBC (“RStudio”) [@rstudio2020]. R code and data are available as a research compendium for generating this paper using the following link: https://github.com/sheilarm23/rrcompendium.git

@@

fastaRF <- read.dna(here(file="analysis/data/raw_data/Sicydium_COI_newRF.fasta"), format = "fasta")

@@

```{r tree, include=TRUE, results = 'hide', fig.show='hide', warning = FALSE}

@@

set.seed(123)
fit = suppressWarnings(optim.pml(fitStart, rearrangement = "stochastic", optGamma=TRUE, optInv=FALSE, model="HKY"))     # seeking a maximum likelihood solution
set.seed(123)    # getting the same tree topology each time the code runs 
bs = bootstrap.pml(fit, bs=100, optNni=TRUE, multicore=TRUE, mc.cores = 4)

# saving the tree in the folder 'analysis/figures/'
bsTree <- plotBS(midpoint(fit$tree), bs, p = 75, type="p", bs.col = "blue")     # fit$tree is where the optimal tree is stored
optimalTree <- plotBS(bsTree, type = "p", cex = 0.2, font = 1)
# export tree with bootstrap values as node labels
write.tree(optimalTree, here(file="analysis/data/derived_data/Sicydium_ML_tree.tre"))

```{r loglikelihood, include = TRUE, results='hide'}
# this result wont be displayed in the final document
loglikelihood <- round(-1*(as.numeric(fit[1])), 2)

```{r editingTree}

# Editing and visualizing the tree to represent --- library: ggtree
tree <- read.tree(here(file="analysis/data/derived_data/Sicydium_ML_tree.tre"))

# checking that the tree is rooted
rootedTree <- is.rooted(tree)   # package ape. showing the tree root: geom_rootpoint(color="pink", size=5) 
# getting the node number for each main clade. This info is important because those #s are the one needed to edit each main clade/species
set.seed(123)
nodeNumber <- ggtree(tree) + geom_text(aes(label=node), hjust=-.5, col = "blue", size = 1) # looks horrible but it gives the clade number

@@

The final alignment comprised 209 sequences, each one of 639 base pairs. Figure \@ref(fig:Sicydium_ML_tree) depicts the ML tree (-lnL = `r loglikelihood`) showing the relationships among all COI sequences identified as *Sicydium* spp. available in GenBank. Six monophyletic clades matching known *Sicydium* species were recovered with strong support. The phylogenetic reconstruction shows that the six samples from Cuba, highlighted in green, nested within *Sicydium buscki* Evermann & Clark, 1906.

@@

@@

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

@@

We thank the Habitat production team for their help during the field work. Sequencing was partially supported by a Smithsonian Tropical Research Institution Short-term Fellowship to SRM. In this sense, we thank advisement provided by Harilaos Lessios and the invaluable lab assistance of Eyda Gómez, Ligia and Axel Calderón. Additional lab work was also done with advice of Diego Elías. R code was developed with help of Tad Dallas, Heru Handika, and the community of Stack Overflow.
